---
title: "The Immune Response of Gasterosteus aculeatus to Schistocephalus solidus and Subsequent Response to Diplostomum pseudospathaceum"
output: html_document
date: "2024-05-16"
---

```{r read-data, include=FALSE}
# READ DATA ----

source("script/stickleback.R")

```


## Introduction
The overall study is focused on immune modulation, the idea that a factor can affect the immune response of an organism.

Parasitism has a number of behavioural, physiological and immunological effects on three-spined Sticklebacks. 

The biological relationship between sticklebacks and diplo can be described by the biological phenomenon of helminth parasitism. This involves modulation of the hosts immune system upon infection with the parasite which enables in to adapt and respond accordingly (2).

The aim of this study was to assess the impacts of infection of parasite A () on the likelihood the stickleback would be infected with parasite B (). Three 'treatments' of fish were studied; high-growth parasite A strain infection, low-growth parasite A strain infection and a group presently uninfected but for which the history is unknown. The control group consisted of fish whomst had never been infected with parasite A. The four groups were exposed to diplo, dissected, and their contents of diplo counted with the aim to discern whether exposure to parasite A had a positive or negative influence on the infection of the host with parasite B. The aims were to identify how current A infection affects B infection and how previous A infection infects B infection. The hypothesis being that the low-growth or cleared infection of parasite A would offer some level of protection from parasite B over the control.
vaccines - immunity effects positive, what are stickleback used for ... could be helpful


## Analysis
The problem addressed here is whether there is an immune benefit to being infected with a less damaging parasite and what the cost-benefit analysis is to the fish.

The data was collected from ... they measured parasite A infection in strains of growth; high, low and unifected. They also measured length of the fish pre exposure, age, sex, total diplo, diplo log and diplo in each eye. In this analysis we were interested in the overall diplo infection and its potential impacts on immunity. We also considered the disadvatages of parasite A such as energetic effects of parasitism thus studying length and also looked at potential differences between sexes as energetic and immunological differences may affect them differently. For example, one paper found fish infected with parasite A were slower and less aligned than those without (1).

Linear model 1:
A linear model was run to analyse diplo intensity as a function of explanatory variable treatment. Looking at whether the differences between the control and the treatments were significant, Infected LG crossed zero.
An ANOVA was used to compare diplo intensity with the predictor variable of treatment. control used as the basis At 95% confidence intervals infected HG had 0.46 higher diplo intensity (CI 0.26 and 0.65), infected LG have a lower intensity but the CIs cross zero so not significant (CI -0.48 and 0.06), uninfected was -0.34 less intensity on average (CI -0.57 and -0.11) which was also significant at a p-value of <0.05. (write up - lecture 3, slide 8). 22 % of the difference in diplo intensity is associated with which treatment (parasite A) was used, with an adjusted R-squared of 0.21. With a p-value of 1.94e-11 overall we can reject the null hypothesis that this would occur by chance.

### Linear modeling:
I used an ordinary least squares method to analyse diplo intensity with parasite A treatments as categorical predictors. The linear model assumption of normality of residues was violated, therefore a square-root data transformation was undertaken to improve the fit of the model.
Sticklebacks with a high-growth parasite A strain infection had a mean increased diplo intensity of a 0.17(95%CI;0.09:0.26) increase from the control sticklebacks with a mean 1.33 diplo intensity (95%CI;1.27:1.39). This was statistically significant(t~214~ = 4.04, *p* = 0.0000732). 
Sticklebacks with a low-growth parasite A strain infection on average had a mean decreased diplo intensity of -0.08(95%CI;-0.19:0.04) from the control sticklebacks which was not statistically significant (t~214~ = -1.27, *p* = 0.21).
Finally, comparison between previously infected sticklebacks and never infected sticklebacks revealed a mean decrease of -0.16(95% CI;-0.25:-0.06) diplo intensity of the uninfected from the control which was statistically significant (t~214~ = -3.08, *p* = 0.002).
The R^2^ for the model was 0.19, with an adjusted R^2^ of 0.18, implying 18% of the variance is explained by the model.
we can therefore reject the null hypothesis

I also used an ordinary least squares method to analyse the impact of treatment (parasite A) as the predictor variable as its effect on length. The model violated the assumption of normality of residues to a limited extent and data transformation did not improve this outcome. sticklebacks with a high-growth infection strain of parasite A, had a mean increase in length of 0.02 mm(95%CI;-1.24:1.28) increase from control sticklebacks (mean = 41.4 mm(95% CI;40.53:42.27)), however this was not statistically significant (t~160~ = 0.03, *p* = 0.98). Sticklebacks with a low-growth parasite A strain infection have a mean increase in length of 0.98 mm(95%CI;-0.55:2.52) which is not statistically significant (t~160~ = 1.27, *p* = 0.21). Finally, the uninfected sticklebacks had a mean increase in length of 0.15 mm(95% CI;-1.27:1.56) which is also not statistically significant (t~160~ = 0.20, *p* = 0.84). The model had an R^2^ of 0.01 and an adjusted R^2^ of -0.01 confering insignificance of the explanatory variables. Not only are there no statistically significant differences, the biological implication of an average of 0.02 mm difference, for example, is limited.
we can therefore accept the null hypothesis.

A t-test was undertaken to determine whether some of the variance in stickleback length could be determined by sex as the predictor variable. The model identified an mean of -1.05 mm(95%CI;-2.03:-0.07) increase in length of female stickleback, statistically significant at (t~161~ = -2.12, *p* = 0.04). The R^2^ of 0.027 and adjusted R^2^ of 0.02, states that just 2 % of the variance is explained by the model. 
we can therefore reject the null hypothesis

Rversion
Rpackages

## Results and Discussion

### Parasite A Infection Impact on Parasite B
Hypothesis 1: The impact of a present A infection will have a negative consequence for the immune system therefore infection of parasite B will be higher.
Hypothesis 2: Previous potential infection of parasite A will have a beneficial consequence for parasite B protection.


```{r plot-1, eval = TRUE, echo = FALSE}

plot_1

```

1 Plot of LG and HG against control

2 Plot of Uninfected against control - shows is superior to having never been infected

visual model summaries and model tables 
3 summary figure (red, orange, yellow)

model findings

overall this shows that there is potential benefit to previous infetion. However, the downside of the uninfected data is we do not know how many of them were infected we just see a sample of the poplation however seeing as it is very different from the control we can assume a great proportion had previous contact with parasite A. Other potential considerations are whether there is a disadvantage to having parasite A energetically - which could be measured by size and on sex, are one sex more affected - what are the reproductive impacts of this? 

### biological impact of parasite A on length and sex

4 plot of length per treatment 
combine
5 plot of length per treatment broken down into sex

model findings


## Conclusion 
yes there are benefits but at what cost?
cost more severe to a specific group?
the next steps



