---
title: "The Response of *Gasterosteus aculeatus* to Historical Infection with *Schistocephalus solidus* Provides Protectective Immunity Against Subsequent Infection with *Diplostomum pseudospathaceum*"
output: html_document
date: "2024-05-16"
---

```{r read-data, include=FALSE}
# READ DATA ----

source("script/stickleback.R")

```


## Introduction
Sticklebacks are small freshwater and oceanic fish, characterised by their abundance of small spines. *Gasterosteus aculeatus*, three-spined sticklebacks, in conjunction with a species of tapeworm (*Schistocephalus solidus*), comprise a robust model system for the study host-parasite relationships (1). The biological phenomenon of helminth parasitism is exhibited by this system, the capacity for immune modulation to be enacted by a parasite on its host, altering future susceptibility to parasite infection (2). *Diplostomum pseudospathaceum*, a second parasitic worm species - trematode, has been identified in a number of bird and fish species, including sticklebacks. In this study we aim to identify how infection with parasite A (*Schistocephalus solidus*), may bolster the immune system of sticklebacks such that they exhibit a level of resistance to infection with parasite B (*Diplostomum pseudospathaceum*). Three levels of parasite A infection were measured; high-growth, low-growth and presently uninfected (historical infection unknown). The control group were never exposed to parasite A. The four groups were exposed to parasite B with infection intensity measured. The primary hypothesis is that a historical, cleared infection of parasite A will afford the uninfected group greater protection from parasite B than the control group. Conversely, current infection with parasite A will be detrimental to the immune system of the sticklebacks, rendering them more susceptible to infection with parasite B than the control group. The secondary hypothesis concerns the energetic costs of parasite A on sticklebacks, which may undermine its potential immunological benefits. It is expected that stickbacks exposed to high-growth parasite A will have reduced body length from the control group, perhaps affecting male sticklebacks more severely due to the energetic cost saved by female sitcklebacks unable to reproduce due to parasitic load.


## Analysis
Initially, I aimed to establish whether there was an immunological advantage to infection of the host with a helminth parasite. To address this question, parasite A (Control, Infected HG, Infected LG and Uninfected) was compared with the log of parasite B intensity (right eye + left eye + 1). I used an ordinary least squares method to analyse parasite B intensity with parasite A treatment levels as the categorical predictor, against the control. This model violated the assumption of normality of residues, thus a square-root data transformation was undertaken to improve the fit of the model to the data. 

To establish the biological cost of parasite A on sticklebacks, each treatment level was compared with initial stickleback length (mm), then subdivided into length per sex to compare energetic cost. An ordinary least squares method was employed to analyse treatment (parasite A) as the predictor variable with length (mm). There was moderate violation of linear model assumptions normality of residues, however, data transformation did not improve this outcome, therefore, careful consideration with which this model is applied to the wider population, must be taken. A t-test was performed to directly compare the catergorical variable of sex with the continuous variable of length, the establish biological connection.

R version
R packages

### Linear modeling:
most of this should be R&D
I used an ordinary least squares method to analyse diplo intensity with parasite A treatments as categorical predictors. The linear model assumption of normality of residues was violated, therefore a square-root data transformation was undertaken to improve the fit of the model.
Sticklebacks with a high-growth parasite A strain infection had a mean increased diplo intensity of a 0.17(95%CI;0.09:0.26) increase from the control sticklebacks with a mean 1.33 diplo intensity (95%CI;1.27:1.39). This was statistically significant(t~214~ = 4.04, *p* = 0.0000732). 
Sticklebacks with a low-growth parasite A strain infection on average had a mean decreased diplo intensity of -0.08(95%CI;-0.19:0.04) from the control sticklebacks which was not statistically significant (t~214~ = -1.27, *p* = 0.21).
Finally, comparison between previously infected sticklebacks and never infected sticklebacks revealed a mean decrease of -0.16(95% CI;-0.25:-0.06) diplo intensity of the uninfected from the control which was statistically significant (t~214~ = -3.08, *p* = 0.002).
The R^2^ for the model was 0.19, with an adjusted R^2^ of 0.18, implying 18% of the variance is explained by the model.
we can therefore reject the null hypothesis

I also used an ordinary least squares method to analyse the impact of treatment (parasite A) as the predictor variable as its effect on length. The model violated the assumption of normality of residues to a limited extent and data transformation did not improve this outcome. sticklebacks with a high-growth infection strain of parasite A, had a mean increase in length of 0.02 mm(95%CI;-1.24:1.28) increase from control sticklebacks (mean = 41.4 mm(95% CI;40.53:42.27)), however this was not statistically significant (t~160~ = 0.03, *p* = 0.98). Sticklebacks with a low-growth parasite A strain infection have a mean increase in length of 0.98 mm(95%CI;-0.55:2.52) which is not statistically significant (t~160~ = 1.27, *p* = 0.21). Finally, the uninfected sticklebacks had a mean increase in length of 0.15 mm(95% CI;-1.27:1.56) which is also not statistically significant (t~160~ = 0.20, *p* = 0.84). The model had an R^2^ of 0.01 and an adjusted R^2^ of -0.01 confering insignificance of the explanatory variables. Not only are there no statistically significant differences, the biological implication of an average of 0.02 mm difference, for example, is limited.
we can therefore accept the null hypothesis.

A t-test was undertaken to determine whether some of the variance in stickleback length could be determined by sex as the predictor variable. The model identified an mean of -1.05 mm(95%CI;-2.03:-0.07) increase in length of female stickleback, statistically significant at (t~161~ = -2.12, *p* = 0.04). The R^2^ of 0.027 and adjusted R^2^ of 0.02, states that just 2 % of the variance is explained by the model. 
we can therefore reject the null hypothesis

Rversion
Rpackages

## Results and Discussion

### Immunological impact of *Schistocephalus solidus* on *Gasterosteus aculeatus* susceptibility to *Diplostomum pseudospathaceum*
Hypothesis 1: Historical infection of sticklebacks with parasite A will be of immunological advantage. 
what plot one shows (reference plot 1)

```{r plot-1, eval = TRUE, echo = FALSE}

plot_1

```
Figure legend:

stats (using example)
other paper findings
link to next test (length and sex)


Hypothesis 2: Current infection of sticklebacks with parasite A will be of immunological disadvantage. 
what plot 2 shows (reference plot 2)
summary and ditribution

```{r mean-table, eval = TRUE, echo = FALSE}

mean_table

```

### Biological impact of *Schistocephalus solidus* on *Gasterosteus aculeatus*, measured in length and sex
Hypothesis 1: Infection of high-growth parasite A will decrease size of sticklebacks
Hypothesis 2: Energetic costs of parasite A infection will impact male sticklebacks more severely

what does the pot show

```{r plot-3, eval = TRUE, echo = FALSE}

plot_3

```

stats (using example)
other paper findings
limitations / further study - look at word docu


## Conclusion 
yes there are benefits but at what cost?
cost more severe to a specific group?
the next steps
overall this shows that there is potential benefit to previous infetion. However, the downside of the uninfected data is we do not know how many of them were infected we just see a sample of the poplation however seeing as it is very different from the control we can assume a great proportion had previous contact with parasite A. Other potential considerations are whether there is a disadvantage to having parasite A energetically - which could be measured by size and on sex, are one sex more affected - what are the reproductive impacts of this? 


